Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-18 ieee_paper.tex
small changes.

The core of the technique is the use of swizzling and permutation
features of the vectors in the Intel MIC architecture. A vector of 512
bits is composed of 4 lanes (sometimes called channels) of 128 bits
which are made of 4 elements of 32 bits. A swizzling operation allows
to reorder the elements within each lane. While a permutation allows to
change the orders of the lanes. Figure~\ref{fig:swizzling_permutation}
illustrates these two possibilities. Both permutation and swizzling
support common reordering and broadcast. In the code, the ``permute''
function needs to use the {\tt \_mm512\_permute4f128\_epi32}
instruction which is only defined on integers and therefore typecast
are necessary to convert to a vector to/from the integer type.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/swizzling_lanes.pdf}
  \caption{Swizzling enables to reorder or broadcast elements within
    each 128 bit group. Permutation enables to reorder or broadcast
    whole 128 bit groups.}
  \label{fig:swizzling_permutation}
\end{figure}

All the operations that loads data from the memory into a register
allows to transfer 512 bits, so 16 floats at a time. A single nonzero
in the matrix will cause 16 operations to be performed, but it only
uses 4 floats from the matrix and 4 floats from the {\tt x} vector. So
we need to load at once the data for four nonzeros in the matrix,
representing 16 floats from the matrix and 16 floats from the vectors
at once. (The number of nonzeros on a row in our application is
always a multiple of 4, if it was, we would need to add some explicit
zeroes in the data structure to pad it.)

The values from the nonzero element in the matrix are the simplest to
load. One can load the 16 floating point values for four nonzeroes of
the matrix at once. The 4 floats coming from each matrix four 4
consecutive nonzeros are naturally grouped within each lane of the
SIMD vector. One can use a swizzle operation to broadcast the 4
elements of a single matrix to take the whole vector. (The broadcast of 
$A_h^1,A_h^2,A_h^3,A_h^4$ is shown in Figure~\ref{code:mat_mul}.)

Loading the entries from the {\tt x} vector is a little bit more
complicated. First the columns pointers of 4 different nonzeros are
loaded in a vector register and the values are distributed one per
lane of the vector register using an {\tt unpack}
operation\footnote{Notice that there are two calls to unpack for the
  LSB and MSB. Both are mandatory even if one is known not to be
  useful according the documentation of the hardware}. (The column
pointers are named $j,k,l,m$ in Figure~\ref{code:mat_mul}.) They
replicated so that each fill a vector lane using a swizzle
operation. Then each value is multiplied by 4 (because there are 4
vectors) and the offsets $(3,2,1,0)$ is added each lane to obtain the
correct index of the elements of the {\tt x} vector within each lane
using a Fused Multiply-Add operation. Then a {\tt gather} operation is
performed bring the 16 entries of the {\tt x} vector into the SIMD
register. At this point we have a SIMD register where each lane is the
four vector entries for a nonzero element. Using a broadcast
permutation, one can replicate an entry of the four vectors accross
each lane. (The broadcast of $x_m^1,x_m^2,x_m^3,x_m^4$ is shown in
Figure~\ref{code:mat_mul}.)

Now that $A_h^1,A_h^2,A_h^3,A_h^4$ and $x_m^1,x_m^2,x_m^3,x_m^4$ are
properly laid out in the vectors, a partial value for $y_i$ can be
computed. Once all the nonzeros of the row have been processed, the
exact $y_i$ value can be sent to memory using a the No Read No Global
Order hint to optimize traffic on the memory bus.

>>>>

The core of the technique is the use of swizzling and permutation
features of the vectors in the Intel MIC architecture. A vector of 512
bits is composed of four lanes (sometimes called channels) of 128 bits, 
which are made of four segments of 32 bits. A swizzling operation allows
the reorder of the elements within each lane, while a permutation allows 
changes the order of entire lanes. 
Figure~\ref{fig:swizzling_permutation}
illustrates these two possibilities. Both permutation and swizzling
support common reordering and broadcast. In the code, the ``permute''
function needs to use the {\tt \_mm512\_permute4f128\_epi32}
instruction which is only defined on integers and therefore in place typecasts
are necessary to convert to a vector to/from the integer type.
\todo{I added a comment stating we preferred swizzling.}
(Note that we prefer swizzling to permutation if possible because we believe 
it is cheaper. Many instructions have swizzling capability, while permuation
is its own instruction.} \todo{reference?}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/swizzling_lanes.pdf}
  \caption{Swizzling enables the reorder or broadcast of elements within
    each 128 bit group. Permutation enables to reorder or broadcast
    whole 128 bit groups. All lanes must be sizzeled identically.}
  \label{fig:swizzling_permutation}
\end{figure}
\todo{I added to the caption.}

Every operation that loads data from memory into a register
transfers 512 bits, or 16 floats at a time. A single nonzero
in the matrix will cause 16 operations to be performed, but it only
uses 4 floats from the matrix and 4 floats from the {\tt x} vector. So
we would like to execute a single instruction to load the data for four 
nonzeros in the matrix, representing 16 floats from the matrix and 16 floats from the vectors
at once. (The number of nonzeros on a row in our application is
always a multiple of 4; if it weren't, we would add some explicit
zeroes in the data structure for padding.)
\todo{ reread paragraph above. single instruction to load 16 floats. Confused.}

The values from the nonzero elements in the matrix are the simplest to
load. One can load the 16 floating point values for four nonzeroes of
the matrix at once. The four floats coming from each matrix four 4
\todo{something wrong with sentence I am in. Did you discuss matrix ordering?}
consecutive nonzeros are naturally grouped within each lane of the
SIMD vector. One can use a swizzle operation to broadcast the 4
elements of a single matrix to take the whole vector. (The broadcast of 
$A_h^1,A_h^2,A_h^3,A_h^4$ is shown in Figure~\ref{code:mat_mul}.)

Loading the entries from the {\tt x} vector is a little bit more
involved. First, the columns pointers of four different nonzeros are
loaded into a vector register and the values are distributed one per
lane of the vector register using an {\tt unpack}
operation\footnote{Notice that there are two calls to unpack for the
  LSB and MSB. Both are mandatory even if one is known not to be
  useful according the documentation of the hardware}. (The column
pointers are named $j,k,l,m$ in Figure~\ref{code:mat_mul}.) They
are replicated so that each value fills an single vector lane 
using a swizzle operation. Then each value is multiplied by 4 
(because there are 4 vectors) 
and the offset vector $(3,2,1,0)$ is added to each lane to obtain the
correct index of the elements of the {\tt x} vector within each lane
using a Fused Multiply-Add operation. Then a {\tt gather} operation is
performed to bring the 16 entries of the {\tt x} vector into the SIMD
register. At this point we have a SIMD register where each lane is the
four vector entries for a nonzero element. Using a broadcast
permutation, one can replicate an entry of the four vectors across
each lane. (The broadcast of $x_m^1,x_m^2,x_m^3,x_m^4$ is shown in
Figure~\ref{code:mat_mul}.)

Now that $A_h^1,A_h^2,A_h^3,A_h^4$ and $x_m^1,x_m^2,x_m^3,x_m^4$ are
properly laid out in the vectors, a partial value for $y_i$ can be
computed. Once all the nonzeros of the row have been processed, the
exact $y_i$ value can be sent to memory using a the No Read No Global
Order hint to optimize write traffic on the memory bus.
\todo{ added "write" to traffic }
